---
title: "Double ML & Heterogeneous Treatment Effects"
author: "MD SAMIUL HAUQUE CHOWDHURY"
format:
  pdf:
    toc: true           
    toc-depth: 2        
    number-sections: true
editor: visual
---

## 1. **Double ML & Heterogeneous Treatment Effects**

### Data Set Description

```{r}
df <- read.csv("data_fitness_het.csv")

head(df, 5)
```

### Partial Linear Model

```{r}
# Load required packages
library(hdm)
library(mlr3)
library(mlr3learners)
library(tidyverse)

set.seed(1234) # for replicability

# Outcome and Treatment
Y <- df$engagement
D <- df$notification_sent

# Create folds for cross-fitting
N <- nrow(df)
nfolds <- 5
fold <- sample(1:nfolds, N, replace = TRUE)

# Initialize nuisance parameter vectors
ehat <- rep(NA, N)      # Propensity score
muhat <- rep(NA, N)     # Outcome model prediction

# Define task and learner for the treatment model (propensity score)
task_e <- as_task_classif(df |> select(-engagement), target = "notification_sent")
lrnr_e <- lrn("classif.ranger", predict_type = "prob")

# Define task and learner for the outcome model
task_mu <- as_task_regr(df |> select(-notification_sent), target = "engagement")
lrnr_mu <- lrn("regr.ranger", predict_type = "response")

# Cross-fitting to estimate nuisance functions
for (i in 1:nfolds) {
  train_idx <- which(fold != i)
  test_idx  <- which(fold == i)

  # Fit propensity score model
  lrnr_e$train(task_e, row_ids = train_idx)
  ehat[test_idx] <- lrnr_e$predict(task_e, row_ids = test_idx)$prob[, 2]

  # Fit outcome model
  lrnr_mu$train(task_mu, row_ids = train_idx)
  muhat[test_idx] <- lrnr_mu$predict(task_mu, row_ids = test_idx)$response
}

# Compute residuals
res_y <- Y - muhat
ATE   <- D - ehat   


# Final residual-on-residual regression
# The coefficient on ATE is the estimated Average Treatment Effect (ATE)
dml_result <- lm(res_y ~ ATE)
summary(dml_result)




```

We can see a **5.578minutes** of average engagement increase after sending notifications to the users. We can see the P-value and standard error is pretty small hence the result is pretty significant.

## 2. **ATE of Notification Sent via Double ML (AIPW Model)**

```{r}
library(hdm)
library(mlr3)
library(mlr3learners)
library(tidyverse)
library(dplyr)

set.seed(1234)

# --- Define target & treatment ---
Y_var <- "engagement"
T_var <- "notification_sent"

# --- Make sure X_vars is a character vector of column NAMES ---
X_vars <- setdiff(names(df), c(Y_var, T_var, "row_id"))

# Cross-fitting setup
N      <- nrow(df)
nfolds <- 5
fold   <- sample(1:nfolds, N, replace = TRUE)

# Prepare storage for nuisance estimates
ehat    <- numeric(N)
muhat0  <- numeric(N)
muhat1  <- numeric(N)

df$row_id <- seq_len(N)
task_e    <- as_task_classif(df %>% select(all_of(c(X_vars, T_var))), target = T_var)
task_mu   <- as_task_regr(   df %>% select(all_of(c(X_vars, Y_var))), target = Y_var)

lrnr_e  <- lrn("classif.ranger",  predict_type = "prob",     num.trees = 500)
lrnr_mu <- lrn("regr.ranger",      predict_type = "response", num.trees = 500)

# --- 3. Cross-fitting of nuisance parameters ---
for (i in seq_len(nfolds)) {
  train_idx   <- which(fold != i)
  predict_idx <- which(fold == i)

  train_ids   <- df$row_id[train_idx]
  pred_ids    <- df$row_id[predict_idx]

  # 1) Propensity score e(x)
  lrnr_e$train(task_e, row_ids = train_ids)
  ehat[predict_idx] <- lrnr_e$predict(task_e, row_ids = pred_ids)$prob[, 2]

  # 2) Outcome mu(0, x)
  lrnr_mu$train(
    task_mu,
    row_ids = intersect(train_ids, which(df[[T_var]] == 0))
  )
  muhat0[predict_idx] <- lrnr_mu$predict(task_mu, row_ids = pred_ids)$response

  # 3) Outcome mu(1, x)
  lrnr_mu$train(
    task_mu,
    row_ids = intersect(train_ids, which(df[[T_var]] == 1))
  )
  muhat1[predict_idx] <- lrnr_mu$predict(task_mu, row_ids = pred_ids)$response
}

# --- DML-AIPW Estimation ---
# Trim propensity scores to avoid extreme values
eps <- 0.01
ehat <- pmin(pmax(ehat, eps), 1 - eps)

psi_a_aipw <- rep(-1, N)
psi_b_aipw <- (muhat1 - muhat0) +
  df[[T_var]] * (df[[Y_var]] - muhat1) / ehat -
  (1 - df[[T_var]]) * (df[[Y_var]] - muhat0) / (1 - ehat)

# Point estimate
theta_aipw <- -sum(psi_b_aipw) / sum(psi_a_aipw)

# Influence function for ATE (psi_i = psi_b_i - theta * psi_a_i)
psi_if <- psi_b_aipw - theta_aipw * psi_a_aipw

# Standard error: sample standard deviation of influence function / sqrt(N)
se_aipw <- sd(psi_if) / sqrt(N)

t_aipw <- theta_aipw / se_aipw
p_aipw <- 2 * pt(abs(t_aipw), df = N - 1, lower.tail = FALSE)

# --- AIPW Results (4 decimal places, non-scientific) ---
cat("AIPW Results::\n")
cat(sprintf("Estimate: %.4f\n", theta_aipw))
cat(sprintf("Std_Error: %.4f\n", se_aipw))
cat(sprintf("t_value: %.4f\n", t_aipw))
cat(sprintf("p_value: %.4f\n", p_aipw))



```

After AIPW modeling sending notification to users increases 5.55 minutes average engagement with a higher t_value and pretty lower standard error. Interestingly the p_value is 0, overall makes our effect robust.

## 3. **Verification with `DoubleML` Package**

```{r}
# Load package
library(DoubleML)

# Specify data object
fit_life_dml <- DoubleMLData$new(
  data = as.data.frame(df),
  y_col = "notification_sent", 
  d_cols = "engagement"
  )
```

```{r}
# Load necessary packages
library(DoubleML)
library(mlr3)
library(mlr3learners)
library(DoubleML)

df <- read.csv("data_fitness_het.csv")

fit_life_dml <- DoubleMLData$new(
  data = df,
  y_col = "engagement",             # outcome
  d_cols = "notification_sent"      # treatment 
)

# Step 2: Specify learners for outcome and treatment models
lrnr_m <- lrn("regr.ranger", num.trees = 100)  # ml_l: outcome model (Y ~ X)
lrnr_e <- lrn("regr.ranger", num.trees = 100)  # ml_m: treatment model (D ~ X)

# Step 3: Create DML-PLR estimator object
dml_plr_obj <- DoubleMLPLR$new(
  data = fit_life_dml,
  ml_l = lrnr_m,                  # outcome model
  ml_m = lrnr_e,                  # treatment model
  apply_cross_fitting = TRUE,
  n_folds = 5                    # 5-fold cross-fitting
)

# Step 4: Fit the model and summarize results
dml_plr_obj$fit()
dml_plr_obj$summary()

```

```{r}
library(DoubleML)
library(mlr3)
library(mlr3learners)

library(DoubleML)

df <- read.csv("data_fitness_het.csv")

fit_life_dml <- DoubleMLData$new(
  data = df,
  y_col = "engagement",             # outcome
  d_cols = "notification_sent"      # treatment 
)

# Step 2: Define learners
lrnr_g <- lrn("regr.ranger", num.trees = 100)     # Outcome model
lrnr_m <- lrn("classif.ranger", num.trees = 100, predict_type = "prob")  # Propensity score model

# Step 3: DoubleMLIRM model
dml_irm_obj <- DoubleMLIRM$new(
  data = fit_life_dml,
  ml_g = lrnr_g,
  ml_m = lrnr_m,
  apply_cross_fitting = TRUE,
  n_folds = 5
)

# Step 4: Fit and summarize
dml_irm_obj$fit()
dml_irm_obj$summary()

```

### Visualization

```{r}
# Load libraries
library(tidyverse)
library(ggplot2)
library(broom)

# --- Manual PLR results (Task 1) 
plr_est <- coef(dml_result)["ATE"]
plr_se <- summary(dml_result)$coefficients["ATE", "Std. Error"]

# --- Manual AIPW results (Task 2) 
aipw_est <- theta_aipw  
aipw_se <- se_aipw

# --- DoubleML PLR results 
dml_plr_est <- dml_plr_obj$coef
dml_plr_se  <- dml_plr_obj$se

# --- DoubleML AIPW results (from IRM model) 
dml_aipw_est <- dml_irm_obj$coef
dml_aipw_se  <- dml_irm_obj$se

# --- Table of all estimates and standard errors 
tbl_coef <- tibble(
  estimator = c(
    "PLR (Manual)",
    "AIPW (Manual)",
    "PLR (DoubleML)",
    "AIPW (DoubleML)"
  ),
  coef = c(
    plr_est,
    aipw_est,
    dml_plr_est,
    dml_aipw_est
  ),
  se = c(
    plr_se,
    aipw_se,
    dml_plr_se,
    dml_aipw_se
  )
)

# --- Plot ---
ggplot(tbl_coef, aes(x = estimator, y = coef, ymin = coef - 1.96 * se, ymax = coef + 1.96 * se)) +
  geom_point(size = 2, color = "steelblue") +
  geom_errorbar(width = 0.2, color = "darkred") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  ylab("Coef") +
  xlab("Estimator") + theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
  expand_limits(y = range(tbl_coef$coef + 2 * tbl_coef$se, tbl_coef$coef - 2 * tbl_coef$se))


print(tbl_coef)
```

Based on this plotting we can see all of our 4 computed effect size is almost same.

1.  AIPW is 5.55 for both manual and package estimation. Difference of t_value is 2 but the standard error and p_value(close to 0) is same and significant.
2.  For the partially linear model, the two ATE estimates differ by only 0.02 (5.57 vs. 5.55). Although the t-values differs by about 10 , the standard errors and p-values remain almost identical and indicate similarly robust significance.

## **4. ATE of Notification Opened via DoubleML**

```{r}
# Load required packages
library(DoubleML)
library(mlr3)
library(mlr3learners)
library(data.table)
library(tidyverse)

# Suppress mlr3 messages
lgr::get_logger("mlr3")$set_threshold("warn")

# Convert your dataset to data.table (required by DoubleML)
df_dt <- as.data.table(df)

# Ensure binary variables are numeric (0/1)
df_dt[, notification_opened := as.numeric(as.character(notification_opened))]
df_dt[, pop_up_rem := as.numeric(as.character(pop_up_rem))]

# Select features: all X variables (excluding treatment, instrument, outcome)
x_vars <- setdiff(colnames(df_dt), c("engagement", "notification_opened", "pop_up_rem"))

# Define DoubleML data object
dml_data <- DoubleMLData$new(
  data = df_dt,
  y_col = "engagement",
  d_cols = "notification_opened",
  z_cols = "pop_up_rem",
  x_cols = x_vars
)

# Define learners
lrn_g <- lrn("regr.ranger")
lrn_m <- lrn("classif.ranger", predict_type = "prob")
lrn_r <- lrn("classif.ranger", predict_type = "prob")

# Initialize DoubleMLIIVM object
dml_iivm <- DoubleMLIIVM$new(
  data = dml_data,
  ml_g = lrn_g,
  ml_m = lrn_m,
  ml_r = lrn_r,
  n_folds = 5,
  trimming_threshold = 0.01
)

# Fit the model
set.seed(123)
dml_iivm$fit()

# Print summary
dml_iivm$summary()

```

We found an **ATE of 9.0094** which is pretty large than our previous ATE which is so obvious as our endogenous variable is notifications_opened now. Additionally small Std error and p value makes this calculation more significant.

### IV Manual computation

```{r}

# Extraction from fitted DoubleMLIIVM
psi_a <- dml_iivm$psi_a
psi_b <- dml_iivm$psi_b
n     <- length(psi_a)

# IV estimate 
theta_iv <- -mean(psi_b) / mean(psi_a)

# Efficient score 
psi       <- psi_b - theta_iv * psi_a

# Standard error, t‐stat and p‐value
se_iv <- sd(psi) / sqrt(n)
t_iv  <- theta_iv / se_iv
p_iv  <- 2 * (1 - pnorm(abs(t_iv)))

cat("Manual IV Results:::\n")
cat("ATE:   ", round(theta_iv, 4), "\n")
cat("Std Error:         ", round(se_iv, 4), "\n")
cat("t-value:    ", round(t_iv, 4), "\n")
cat("p-value:    ", format.pval(p_iv, 4), "\n")


```

After extracting psi_a and psi_b we used manual formula and estimated the output that we got previously. We got the same **ATE=9.009minutes.** While ATE is consistent, the SE, P_value is more smaller and t-value is more bigger which makes our estimation significant than the result of doubleml package.

## **5. Heterogeneous Treatment Effects of Notification Sent**

### Estimate group ATEs by gender (M/F) and by tertiles of past_engagement.

```{r}
# Male
male_df <- subset(df, gender == "Male")
male_ate <- lm(engagement ~ pop_up_rem, data = male_df)
cat("Male ATE \n") 
summary(male_ate)

# Female
female_df <- subset(df, gender == "Female")
female_ate <- lm(engagement ~ pop_up_rem, data = female_df)
cat("Female ATE \n") 
summary(female_ate)

##tertilles of past engagement

cat("%-----tertiles of past engagemen----- %\n") 
library(estimatr)

df <- df %>%
  mutate(
    engage_tertile = factor(
      ntile(past_engagement,3),
      levels = 1:3,
      labels = c("Low","Mid","High")
    )
  )

# Interaction regression
gate_engage <- lm_robust(
  engagement ~ pop_up_rem * engage_tertile,
  data = df
)
summary(gate_engage)


```

### Individualized ATEs with a Causal Forest.

```{r}

library(grf)

# 1) Prepare outcome, treatment and covariates
Y <- df$engagement
W <- df$notification_sent

# One-hot encode and drop intercept:
X <- model.matrix(~ . - engagement - notification_sent, data = df)[, -1]

# 2) Fit the causal forest
set.seed(123)
cf_model <- causal_forest(
  X, Y, W,
  num.trees       = 2000,             
  sample.fraction = 0.5,              
  mtry            = floor(ncol(X)/3), 
  min.node.size   = 5,                
  honesty         = TRUE              )

# 3) Estimate the overall ATE
ate_cf <- average_treatment_effect(cf_model, target.sample = "all")
cat("Causal Forest ATE:", round(ate_cf[1], 4),
    " (SE =", round(ate_cf[2], 4), ")\n")

# 4) Predict individual CATEs (IATEs)
tau_hat <- predict(cf_model, newdata = X)$predictions
df$iate_cf <- tau_hat

# 5) Summarize and visualize
summary(tau_hat)
hist(tau_hat,
     breaks = 30,
     main   = "Causal Forest IATEs",
     xlab   = "Estimated Treatment Effect")



```

### Splitting data set by 80/20

```{r}
set.seed(123)              
n     <- nrow(df)          
i_tr  <- sample(n, 0.8*n)  # 80% of indices

Train <- df[i_tr, ]        # 80% training data
Test  <- df[-i_tr, ]       # remaining 20% test data
# print(n)
# print(Train, Test)
```

### Predict Individualized ATEs by implementing the R-Learner as weighted minimization problem.

```{r}

library(grf)

# Specify your covariates
covariates <- c(
  "age","gender","region","device_type",
  "app_version","sessions_per_day","past_engagement",
  "notification_sent","notification_opened"
)

# Build X, W, Y on train
X_train <- model.matrix(~ . - 1, data = Train[, covariates])
W_train <- Train$pop_up_rem
Y_train <- Train$engagement

# 2. Estimate nuisances on Train
# 2a) Propensity e_hat(X)
ps_mod   <- glm(W_train ~ X_train, family = binomial)
e_hat_tr <- predict(ps_mod,   type = "response")

# 2b) Baseline outcome μ_hat(X)
mu_forest  <- regression_forest(X_train, Y_train)
mu_hat_tr  <- predict(mu_forest)$predictions

# 3. Form pseudo‐outcome and weights
pseudo_R <- (Y_train - mu_hat_tr) / (W_train - e_hat_tr)
wt_R     <- (W_train - e_hat_tr)^2

# 4. Fit the R‐Learner via weighted regression forest
rlearner_forest <- regression_forest(
  X_train,
  pseudo_R,
  sample.weights = wt_R
)

# 5. Extract in‐sample IATEs on 80% (for Task 5)
tau_r_train <- predict(rlearner_forest, X_train)$predictions
# 6. Compute and print the overall ATE
ate_r_train <- mean(tau_r_train)
cat("R-Learner In-Sample ATE:", round(ate_r_train, 3), "\n")

# 7. Look at the distribution of IATEs
print(summary(tau_r_train))

```

### Predict Individualized ATEs by implementing the DR-Learner.

```{r}

#  Propensity via logistic regression
prop_mod  <- glm(pop_up_rem ~ age + gender + region + device_type +
                   app_version + sessions_per_day + past_engagement +
                   notification_sent + notification_opened,
                 data = Train, family = binomial)
Train$e_hat <- predict(prop_mod, type = "response")

#  2b) Outcome regressions m0 and m1
m0_mod    <- lm(engagement ~ age + gender + region + device_type +
                  app_version + sessions_per_day + past_engagement +
                  notification_sent + notification_opened,
                data = Train, subset = (pop_up_rem == 0))
m1_mod    <- lm(engagement ~ age + gender + region + device_type +
                  app_version + sessions_per_day + past_engagement +
                  notification_sent + notification_opened,
                data = Train, subset = (pop_up_rem == 1))

Train$mu0_hat <- predict(m0_mod, newdata = Train)
Train$mu1_hat <- predict(m1_mod, newdata = Train)

#  Form the DR pseudo‐outcome
Train$dr_pseudo <- with(Train,
  mu1_hat - mu0_hat
  + (pop_up_rem   * (engagement - mu1_hat)) / e_hat
  - ((1 - pop_up_rem) * (engagement - mu0_hat)) / (1 - e_hat)
)

#  Final CATE model: regress dr_pseudo on X
tau_mod <- lm(dr_pseudo ~ age + gender + region + device_type +
                app_version + sessions_per_day + past_engagement +
                notification_sent + notification_opened,
              data = Train)

#  In-sample IATEs on your 80 %
Train$tau_dr_train <- predict(tau_mod, newdata = Train)

#  Summaries
ate_dr_train <- mean(Train$tau_dr_train)
cat("DR-Learner In-Sample ATE:", round(ate_dr_train,3), "\n")
summary(Train$tau_dr_train)
hist(Train$tau_dr_train,
     main = "DR-Learner In-Sample IATEs",
     xlab = "Estimated CATE (minutes)")
```

### Descriptive statistics with plotting

```{r}
# compute your three IATE vectors
cf   <- tau_hat                # from causal_forest
rl   <- tau_r_train            # from R-Learner
dr   <- Train$tau_dr_train     # from DR-Learner

# set up common x‐range
xlim <- range(cf, rl, dr)

# plot CF
hist(cf, breaks = 30, xlim = xlim, probability = TRUE,
     col = rgb(0, 0, 1, 0.4), border = NA,
     main = "Overlapping IATE Densities", xlab = "IATE")

# add R-Learner
hist(rl, breaks = 30, probability = TRUE, add = TRUE,
     col = rgb(0, 1, 0, 0.4), border = NA)

# add DR-Learner
hist(dr, breaks = 30, probability = TRUE, add = TRUE,
     col = rgb(1, 0, 0, 0.4), border = NA)

legend("topright",
       legend = c("Causal Forest", "R-Learner", "DR-Learner"),
       fill   = c(rgb(0,0,1,0.4),
                  rgb(0,1,0,0.4),
                  rgb(1,0,0,0.4)),
       border = NA)


```

## 6. **HTE Evaluation**

### Causal Forest IATEs with the Rank-weighted Average Treatment Effect (RATE)

### DR-Learner trained above for the rest 20% dataset

```{r}
# 1. Propensity & outcome predictions on Test (no re‐fit)
Test$e_hat   <- predict(prop_mod, newdata = Test, type = "response")
Test$mu0_hat <- predict(m0_mod,  newdata = Test)
Test$mu1_hat <- predict(m1_mod,  newdata = Test)

# 2. Build the DR pseudo‐outcome on Test
Test$dr_pseudo <- with(Test,
  mu1_hat - mu0_hat
  + (pop_up_rem * (engagement - mu1_hat)) / e_hat
  - ((1 - pop_up_rem) * (engagement - mu0_hat)) / (1 - e_hat)
)

# 3. Predict IATEs for the 20% hold-out
Test$tau_dr_test <- predict(tau_mod, newdata = Test)

# 4. Summarize
cat("DR-Learner OOS ATE:", round(mean(Test$tau_dr_test),3), "\n")
summary(Test$tau_dr_test)
hist(Test$tau_dr_test,
     main = "DR-Learner Out-of-Sample IATEs",
     xlab = "Estimated CATE (minutes)")


```

## 7. Optimal policy learning

### Classification tree to learn an optimal treatment policy of whom should be sent a push notification.

```{r}
# Load required packages
library(mlr3)
library(mlr3learners)
library(tidyverse)
library(rpart.plot)
# Get and transform the individual ATEs (pseudo-outcomes)
df$ate_i   <- dml_irm_obj[["psi_b"]]
df$sign    <- sign(df$ate_i)
df$weights <- abs(df$ate_i)

# Optimal Policy Learning with Classification Tree
df_sub <- df %>%
  select(
    sign, weights, age, gender, region, device_type,
    app_version, sessions_per_day, past_engagement,
    notification_sent, notification_opened
  ) %>%
  mutate(
    gender      = factor(gender),
    region      = factor(region),
    device_type = factor(device_type)
  )

task <- as_task_classif(df_sub, target = "sign")
task$set_col_roles("weights", roles = "weights_learner")
# Increase tree complexity to allow splits, including by age
lrnr <- lrn("classif.rpart", cp = 1e-04, minsplit = 10, maxdepth = 5)
lrnr$train(task)
rpart.plot(lrnr$model)

# Compute policy assignment pi = {-1,1}
df_sub$pi <- 2 * (as.numeric(lrnr$predict(task)$response) - 1.5)

# Classification Analysis
covariate_matrix <- model.matrix(
  ~ age + app_version + sessions_per_day + past_engagement + notification_sent + notification_opened
    + gender + region + device_type - 1,
  data = df_sub
)

# Compute CLAN: group-level means of each numeric covariate
CLAN <- rbind(
  No_Reminder = colMeans(covariate_matrix[df_sub$pi == -1, ], na.rm = TRUE),
  Reminder    = colMeans(covariate_matrix[df_sub$pi ==  1, ], na.rm = TRUE)
)

# Transpose for readability and round
CLAN <- round(t(CLAN), 2)
CLAN

```
