---
title: "Assignment-01"
author: "Md Samiul Hauque Chowdhury"
date: "04-05-2025"
format:
  pdf:
    documentclass: article
    number-sections: false
    toc: false
    toc-depth: 1
    code-overflow: wrap
    keep-tex: true
    include-in-header: preamble.tex
    fig-cap-location: top
    tbl-cap-location: top
fontsize: 11pt
geometry: margin=1in
bibliography: citations.bib
link-citations: true
csl: ieee.csl
editor: visual
---

## 1. Dataset and Scenario Description

This dataset is originally from the National Institute of Diabetes and Digestive and Kidney Diseases.[@kaggle2024diabetes]. It includes various health-related attributes of female patients, and a binary outcome indicating the presence or absence of diabetes. Our objective is to find how BMI leads a female person towards diabetes.

**`Key variables used:`**

-   Pregnancies: Number of times pregnant.

-   Glucose: Plasma glucose concentration after a 2-hour oral glucose tolerance test.

-   BloodPressure: Diastolic blood pressure (mm Hg).

-   SkinThickness: Triceps skin fold thickness (mm).

-   Insulin: 2-hour serum insulin (mu U/ml).

-   BMI: Body mass index (computed as weight (kg)/height (m)2).

-   DiabetesPedigreeFunction: Diabetes pedigree function.

-   Age: Age (in years).

-   Outcome: Diagnosis result (1 for diabetic, 0 for non-diabetic)

Unfortunately one important variable is missing here which is commonly refer in case of Diabetes is "Dietary Habbits". Hence with the help of some AI Agent we simulated another column Dietary Habbit which specifies 1=Good, 2=Moderate, 3=Poor.

```{r}
# Load required packages
library(dplyr)
library(readr)

# STEP 1: Load the dataset
df <- read_csv("diabetes.csv")

# STEP 2: Simulate 'DH' (Dietary Health) based on only BMI and Glc
set.seed(42)

simulate_dh <- function(bmi, glc) {
  if (is.na(bmi) | is.na(glc)) {
    return(NA)
  } else if (bmi > 35 | glc > 150) {
    return(sample(c(2, 3), 1, prob = c(0.3, 0.7)))
  } else if (bmi < 25 & glc < 100) {
    return(sample(c(1, 2), 1, prob = c(0.7, 0.3)))
  } else {
    return(sample(c(1, 2, 3), 1, prob = c(0.3, 0.4, 0.3)))
  }
}


# STEP 4: Save the updated dataset
write_csv(df, "diabetes.csv")

# STEP 5: Show the first 5 rows
head(df, 5)



```

### Key Questions:

1.  Does BMI have a causal effect on diabetes for female patients?
2.  How other important variables (Pregnancy, Age, Blood Pressure) influence our T= BMI and Y=diabetes?
3.  What are the significant mediators, con-founders and potential con-founders.

## 2. Developing a Graphical Causal Model (DAG)

```{r}

# Load libraries
library(tidyverse)
library(ggdag)
library(dagitty)
library(broom)

# Adjusted node layout for better spacing (preserved)
node_details <- tribble(
  ~name, ~label, ~x, ~y,
  "DH", "Diet", 2, 10,
  "BMI", "BMI", 1, 6,
  "BP", "BP", 6, 4,
  "Ins", "Insulin", 9, 8,
  "Glc", "Glc", 13, 4,
  "Diabetes", "Diabetes", 15, 7,
  "DPF", "DPF", 7, 10,
  "Age", "Age", 5, 10,
  "PREG", "Pregnancies", 2, 4,
  "ST", "SKinT", 13, 10
)

# Attach readable labels
node_labels <- node_details$label
names(node_labels) <- node_details$name

# Define DAG with treatment and outcome properly set
dag1 <- dagify(
  BMI ~ Age + PREG +DH ,
  BP ~ DH + Age + BMI + PREG,
  Ins ~ BP + BMI + PREG + DPF,
  Glc ~ Ins +DH,
  Diabetes ~ Glc,
  
  ST ~ Age + Ins,
  
  
  
  exposure = "BMI",        
  outcome  = "Diabetes",    
  coords = node_details,
  labels = node_labels
)

# Tidy DAG and automatically assign status (exposure, outcome, latent)
dag_tidy <- dag1 %>%
  tidy_dagitty() %>%
  node_status()

# Define color scheme (orange for exposure, teal for outcome)
status_colors <- c(
  exposure = "#005e73",  # BMI
  outcome  = "#FF7E15",  # Diabetes
  latent   = "grey50"
)

# Final DAG plot
ggplot(dag_tidy, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(aes(color = status)) +
  geom_dag_edges() +
  geom_dag_label_repel(
    aes(label = label, fill = status),
    size          = 2.5,
    seed          = 1234,
    color         = "white",
    fontface      = "bold",
    point.padding = unit(0.6, "lines"),
    box.padding   = unit(0.6, "lines"),
    nudge_y       = 1
  ) +
  scale_color_manual(values = status_colors, na.value = "#00C1D4") +
  scale_fill_manual(values = status_colors, na.value = "#00C1D4") +
  guides(color = "none", fill = "none") +
  theme_dag()




```

### Explanations & Justifications of Important DAG relationship

### **Confounders**

**Age → Blood Pressure, Age → Skin Thickness** and **Age → BMI**,:\
Physiological changes associated with aging—such as arterial stiffening and reduced vascular compliance—lead to progressive increases in systolic blood pressure, particularly isolated systolic hypertension in older adults.[@singh2023bp]

**Pregnancies → BMI, Pregnancies →Blood Pressure and Pregnancies →Insulin**:\
The number of pregnancies can affect body weight and long-term metabolic changes, influencing insulin resistance and blood pressure. Gestational diabetes describes how pregnancy is directly relation to insulin resistance and eventually diabetes. [@slootweg2023maternal]

**Diabetes Pedigree Function (DPF) → Insulin**:\
DPF approximates a person's genetic susceptibility to diabetes. This family/genetic background can influence intermediate physiological variables like insulin sensitivity.

### **Potential Confounders**

**Dietary Habit → BMI** and **Dietary Habit → Blood Pressure**:\
People with unhealthy dietary habits—like eating fatty, salty foods—are more likely to gain weight and develop high blood pressure. Over time, this can lead to insulin resistance and increase the risk of type 2 diabetes. [@li2024hypertension]

### **Mediators**

**BMI → Blood Pressure → Insulin → Glucose → Diabetes**:\
Excess body fat triggers the kidneys and hormones to raise blood pressure, also making the body less responsive to insulin and increasing nervous system activity. This leads to higher blood sugar levels and can cause diabetes. [@kalandarova2024diet] [@hall2015obesity]

**Insulin → Skin Thickness** :\
According to a study, a small amount of insulin can cause increased skin thickness. [@murao2022insulin]

## 3. Causal Discovery

```{r pc-graph, message=FALSE, warning=FALSE, results='hide'}

# Load required libraries
library(tidyverse)
library(pcalg)
library(graph)
library(RBGL)

# STEP 1: Read the cleaned and overwritten dataset
diabetes <- read_csv("diabetes.csv")  # no renaming needed

# STEP 2: Subset variables for DAG learning
vars <- c("Age", "BMI", "BP", "Glc", "Ins", "Out", "DPF", "Preg", "Skin", "DH")
df <- diabetes[, vars]

# STEP 3: Ensure all variables are numeric (for correlation matrix)
df <- df %>% mutate(across(everything(), as.numeric))

# STEP 4: Create sufficient statistics for Gaussian CI test
suffStat <- list(C = cor(df), n = nrow(df))

# STEP 5: Apply the PC algorithm
pc.fit <- pc(
  suffStat = suffStat,
  indepTest = gaussCItest,
  alpha = 0.05,
  labels = colnames(df),
  verbose = TRUE
)

# STEP 6: Plot the learned CPDAG
plot(pc.fit, main = "Simplified Causal Graph")



```

After DAG simplifications a few edges were omitted. A new **bidirectional** edge has been introduced with pregnancy to age. Importantly, a new edge is visible from our **treatment(BMI)** to **outcome(Diabetes)**. Same goes for pregnancy to outcome. In our original DAG a relation between **Diabetes** to DPF(**Diabetes Pedigree Function)** is missing.

However, we can draw a new DAG based on the result given from PcAlgorithm.

### DAG Simplifications

```{r}
# Load libraries
library(tidyverse)
library(ggdag)
library(dagitty)
library(broom)

# Node layout with labels
node_details <- tribble(
  ~name, ~label, ~x, ~y,
  "DH", "Diet", 2, 10,
  "BMI", "BMI", 1, 6,
  "BP", "BP", 6, 4,
  "Ins", "Insulin", 9, 8,
  "Glc", "Glc", 13, 4,
  "Diabetes", "Diabetes", 15, 7,
  "DPF", "DPF", 7, 10,
  "Age", "Age", 5, 10,
  "PREG", "Pregnancies", 2, 4,
  "ST", "SKinT", 13, 10
)

# Attach readable labels
node_labels <- node_details$label
names(node_labels) <- node_details$name

# Define DAG (fixed node names + correct labels/coords)
dagSim <- dagify(
  ST ~ BP + BMI + DPF + Ins + Age,
  Diabetes ~ BMI + PREG + Glc,
  DH ~ BMI  + Glc,
  Glc ~ Age,
  Ins ~ Glc,
  PREG ~ Age,
  BP ~ Age + BMI,
  Age ~ PREG,
  DPF ~ Ins+Diabetes,
  
  exposure = "BMI",
  outcome = "Diabetes",
  
  coords = node_details,
  labels = node_labels
)

# Tidy DAG and mark node roles
dag_tidy <- dagSim %>%
  tidy_dagitty() %>%
  node_status()

# Define colors
status_colors <- c(
  exposure = "#005e73",  # Dark teal for exposure
  outcome  = "#FF7E15",  # Orange for outcome
  latent   = "grey50"
)

# Final plot
ggplot(dag_tidy, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(aes(color = status)) +
  geom_dag_edges() +
  geom_dag_label_repel(
    aes(label = label, fill = status),
    size = 2.5,
    seed = 1234,
    color = "white",
    fontface = "bold",
    point.padding = unit(0.6, "lines"),
    box.padding = unit(0.6, "lines"),
    nudge_y = 1
  ) +
  scale_color_manual(values = status_colors, na.value = "#00C1D4") +
  scale_fill_manual(values = status_colors, na.value = "#00C1D4") +
  guides(color = "none", fill = "none") +
  theme_dag()


```

## 4. Identifications of Causal Effects

In order to do the identifications of causal effects, we must execute backdoor criterion, frontdoor criterion and do -calculus.

### Backdoor Adjustment

So the minimum adjustment set for our initial graph is below:

```{r}
adjustmentSets(dag1, exposure = "BMI", outcome="Diabetes")
```

We must need to condition on Age and pregnancy in order to validate our causal relation from BMI to Diabetes. For example, older individuals tend to have higher BMI and are also more likely to develop diabetes.

**Collider Check**: We must need to careful about collider, as conditioning on ***Insulin*** will open new backdoor paths. Although Insulin here acts a potential collider.

```{r}
isAdjustmentSet(dag1, c("Age", "DH", "PREG", "Ins"), exposure="BMI", outcome="Diabetes")
```

On the other hand, our simplified DAG, minimal adjustment set is:

```{r}
adjustmentSets(dagSim, exposure ="BMI", outcome= "Diabetes")
```

We don't need to condition on **any variable** in case of our simplified DAG. But we must need to be careful about collider and not condition them.

### Frontdoor Adjustment

[**Condition 1**]{.underline}

**Original DAG:**

After blocking all the backdoor path we can see 2 set of mediating path,

1.  BMI → Insulin → Glucose → Diabetes
2.  BMI → BP → Insulin → Glucose → Diabetes

BMI is direclty mediating through mediator variables and causing diabetes **satisfy the 1st condition**.

**Simplified DAG:**

There is a direct edge from BMI to Diabetes. Hence there is no mediator between them and **does not satisfy** the 1st condition. Although theoretically BMI is not a direct reason for diabetes. It must need to have an influential relation with insulin resistance. [@klein2019obesity]

[**Condition 2**]{.underline}

**Original DAG:**

There should not be any unblocked backdoor paths from BMI to Mediators thus second condition does not met.

```{r}
adjustmentSets(dag1, exposure = "BMI", outcome="Glc")
```

```{r}
adjustmentSets(dag1, exposure = "BMI", outcome="Ins")
```

```{r}
adjustmentSets(dag1, exposure = "BMI", outcome="BP")
```

**Simplified DAG:**

```{r}
#adjustmentSets(dagSim, exposure = "BMI", outcome="")
```

Obviously simplified DAG will not satisfy the 2nd condition as there is no mediator.

[**Condition 3**]{.underline}

**Original DAG:**

All backdoor paths from Mediators to Outcome need to be blocked my BMI. Hence 3rd condition does not met.

```{r}
adjustmentSets(dag1, exposure = "Ins", outcome="Diabetes")
```

```{r}
adjustmentSets(dag1, exposure = "BP", outcome="Diabetes")
```

```{r}
adjustmentSets(dag1, exposure = "Glc", outcome="Diabetes")
```

**Simplified DAG:**

There is no mediator here to execute so this DAG failed here as well.

### do-calculus

**Original DAG:**

```{r}
library(dosearch)

dag_do <- "
  Age -> BMI
  PREG -> BMI
  DH -> BMI
  
  DH -> BP
  Age -> BP
  BMI -> BP
  PREG -> BP

  BP -> Ins
  BMI -> Ins
  PREG -> Ins
  DPF -> Ins

  Ins -> Glc
  DH -> Glc

  Glc -> Diabetes

  Age -> ST
  Ins -> ST
"

# All observed variables (assuming no hidden variables for now)
data_str <- "P(Age, PREG, DH, DPF, BMI, BP, Ins, Glc, Diabetes, ST)"

# The query of interest: effect of BMI on Diabetes
query <- "P(Diabetes | do(BMI))"

# Run do-search
res <- dosearch(data_str, query, dag_do)

# Check identifiability
is_identifiable(res)

# If identifiable, see the formula
res$formula

```

```{r}
get_derivation(res)
```

```{r}
cat(res$formula)

```

Expression for original DAG:

$$
 \sum_{\text{Age}, \text{PREG}, \text{DH}, \text{DPF}} \left[ P(\text{Age}, \text{PREG}, \text{DH}, \text{DPF}) \cdot P(\text{Diabetes} \mid \text{BMI}, \text{Age}, \text{PREG}, \text{DH}, \text{DPF}) \right]
$$

**Simplified DAG:**

```{r}
library(dosearch)

# Query and DAG
new_query <- "P(Diabetes | do(BMI))"
dag_code <- "
  Age -> BMI
  PREG -> BMI
  DH -> BMI
  BMI -> BP
  BMI -> Ins
  BP -> Ins
  PREG -> Ins
  DPF -> Ins
  Ins -> Glc
  DH -> Glc
  Glc -> Diabetes
  Age -> ST
  Ins -> ST
"
data_str <- "P(Age, PREG, DH, BMI, BP, Ins, Glc, Diabetes, DPF, ST)"

# Run do-search
res2 <- dosearch(data_str, new_query, dag_code)

# Check identifiability
is_identifiable(res2)

# Correct way to view formula
cat("Derived formula:\n", deparse(res2$formula), sep = "\n")


```

```{r}
get_derivation(res)
```

```{r}
cat(res2$formula)
```

Expression for Simplified DAG:

$$
\sum_{\text{Age}, \text{PREG}, \text{DH}, \text{DPF}} \left( p(\text{Age}, \text{PREG}, \text{DH}, \text{DPF}) \cdot p(\text{Diabetes} \mid \text{BMI}, \text{Age}, \text{PREG}, \text{DH}, \text{DPF}) \right)
$$

## 5. Reflection and critical evaluation

In this assignment, we investigated the causal relationship between BMI and the risk of developing diabetes using a combination of domain knowledge and methods of Causal Discovery. We identified and blocked backdoor paths through proper adjustment (e.g., Age, Dietary Habit, Pregnancy) and confirmed identifiability using the dosearch algorithm. We found that BMI mediates its effect on Diabetes through Blood Pressure, Insulin, and Glucose. Although in our simplified graph we didn't find any adjustment set and interestingly there was a direct (T-\>Y) makes frontdoor calculation more complex. Despite structural variations in DAGs, the causal formula remained consistent.

## 6. Reference

::: {#refs}
:::
